### 2023-01-18 paper.Rmd
Converted paper to Rmarkdown for easier output generation.

---
title: "BumpyMatrix: a Bioconductor package to represent a 2-dimensional array of non-scalar objects"
tags:
  - R
  - Bioconductor
  - bioinformatics
authors:
  - name: Aaron T. L. Lun
    orcid: 0000-0002-3564-4813
    affiliation: 1
affiliations:
  - name: Genentech Inc., South San Francisco, USA
    index: 1
date: 17 January 2022
bibliography: ref.bib
---

```{r, echo=FALSE, results="hide"}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE, collapse=TRUE)
set.seed(100)
```

# Summary

# Statement of need

# Basic usage

Let's mock up some data for a multiplexed FISH experiment.
This consists of a `DataFrame` (Bioconductor's wrapper around the base `data.frame` class) where each row contains the coordinates of a single detected transcript molecule in a particular cell.

```{r}
library(S4Vectors)
df <- DataFrame(
    x=rnorm(10000), y=rnorm(10000), 
    gene=paste0("GENE_", sample(100, 10000, replace=TRUE)),
    cell=paste0("CELL_", sample(20, 10000, replace=TRUE))
)
df
```

We use the `splitAsBumpyMatrix()` function to create a `BumpyDataFrameMatrix` where the rows are genes and the columns are cells.
Each entry of this `BumpyDataFrameMatrix` is itself a `DataFrame` that corresponds to a particular gene/cell combination,
and containing the x/y-coordinates for all transcript molecules of that gene in that cell.
Each `DataFrame` may have zero, one or multiple rows, depending on the cell's expression of the gene.

```{r}
library(BumpyMatrix)
mat <- splitAsBumpyMatrix(df[,c("x", "y")], row=df$gene, column=df$cell)
mat
```

We can now treat `mat` in the same manner as any other matrix-like R object.
For example, we can slice by row and column, equivalent to filtering our original `df` by gene or cell respectively.

```{r}
mat[1:5,]

mat[,c("CELL_5", "CELL_10", "CELL_20")]
```

Extracting a row or column of the `BumpyDataFrameMatrix` will return a list of `DataFrame` objects, which is the 1-dimensional analogue to the atomic vectors obtained from extracting a row/column of an ordinary R matrix.
(More specifically, the return value is a Bioconductor `SplitDataFrameList`, which is used as the underlying data representation for the `BumpyDataFrameMatrix` implementation.)
A single `DataFrame` can be extracted from a `BumpyDataFrameMatrix` or a `SplitDataFrameList` with the usual `[[` operator.

```{r}
mat[1,]

mat[1,1][[1]]
```

The `lengths()` function will return the number of entries in each `DataFrame`.
In this context, the return value is a count matrix that quantifies the expression of each gene in each cell.

```{r}
counts <- lengths(mat)
counts[1:10,1:5]
```

# Compatibility with the `SummarizedExperiment`

Of particular interest is the ability to store `mat` inside Bioconductor's `SummarizedExperiment` data structure.
The `SummarizedExperiment` provides a holistic representation of a dataset that contains all experimental assays, feature- and sample-level annotations as well as additional metadata. 
This representation is convenient as it ensures that any operations are synchronised across all aspects of the dataset, avoiding any account-keeping errors across the lifetime of an analysis workflow.
By storing the `BumpyDataFrameMatrix` inside the `SummarizedExperiment`, we can synchronise the manipulations of the transcript coordinates to the rest of the dataset.

```{r}
library(SummarizedExperiment)
se <- SummarizedExperiment(list(positions = mat, counts = lengths(mat)))
se

# Limiting our analysis to the first 10 cells:
sub.se <- se[,1:10]
dim(assay(sub.se, "counts"))

# Our BumpyDataFrameMatrix is also subject to the same operation:
assay(sub.se, "positions")
```

This compatibility also simplifies the consumption of `BumpyDataFrameMatrix` assays by frameworks that operate on a `SummarizedExperiment`.
For example, the **alabaster** framework provides a mechanism for serializing R/Bioconductor objects into language-agnostic file formats, for use with other programming ecosystems like Python or Javascript.
As **alabaster** already provides a mechanism to serialize `SummarizedExperiment` objects to file, it can be easily extended to handle objects containing `BumpyDataFrameMatrix` objects in its assays.

```{r}
library(alabaster.se) # TODO: change to the alabaster umbrella.

# Saving the SummarizedExperiment into a staging directory.
staging <- tempfile()
dir.create(staging)
info <- stageObject(se, staging, "experiment_1")

# Loading the SummarizedExperiment recovers our BumpyDataFrameMatrix.
reloaded <- loadObject(info, staging)
assay(reloaded, "positions")
```

# Other `BumpyMatrix` subclasses

While we have largely focused on the `BumpyDataFrameMatrix`, the _BumpyMatrix_ package also supports other `BumpyMatrix` subclasses.
For example, the `BumpyNumericMatrix` class implements a matrix-like object where each entry is a numeric vector of any length.
To illustrate, we can generate a `BumpyNumericMatrix` instance by extracting the x-coordinates from `mat`.

```{r}
xpos <- mat[,,"x"]
xpos
```

This can be manipulated in the same manner as a `BumpyDataFrameMatrix`, except that each entry is now a numeric vector instead of a `DataFrame`.

```{r}
# Getting the x-coordinates for the first cell:
xpos[,1]

# Extracting the numeric vector directly:
xpos[1,1][[1]]
```

These `BumpyMatrix` classes can also participate in basic arithmetic, logical and mathematical operations, allowing users to perform complex manipulations in an intuitive manner.
Imagine that we have the centroid for each cell, and we wish to compute the distance of each transcript from its cell's centroid.
We can achieve this by applying the usual formula for a 2-dimensional Euclidean distance on our `BumpyNumericMatrix` objects.

```{r}
# Mocking up a centroid x/y coordinate for each cell.
x_cell <- rnorm(ncol(mat))
y_cell <- rnorm(ncol(mat))

# Computing the Euclidean distance of each transcript from the centroid.
# Transpositions are necessary to ensure that the cell centroid coordinates
# are recycled along the cells, not along the features.
squared_x <- (t(mat[,,"x"]) - x_cell)^2 
squared_y <- (t(mat[,,"y"]) - y_cell)^2
dist <- t(sqrt(squared_x + squared_y))
dist
```

Further imagine that we wish to filter `mat` to remove transcripts that are more than, say, 2 units away from the centroid.
This is possible by converting our `BumpyNumericMatrix` into a `BumpyLogicalMatrix` via the usual logical `<` operation.
The resulting `BumpyLogicalMatrix` can then be used to subset our original `BumpyDataFrameMatrix`, which will apply the filter of the former to the rows of each individual `DataFrame` in the latter.

```{r}
keep <- dist <= 2
keep

# We can use a BumpyLogicalMatrix as a index into the BumpyDataFrameMatrix.
filtered <- mat[keep]
filtered
```

# References

### 2023-01-18 paper.Rmd
Minor rewording here and there.

@@
 
-Let's mock up some data for a multiplexed FISH experiment.
-This consists of a `DataFrame` (Bioconductor's wrapper around the base `data.frame` class) where each row contains the coordinates of a single detected transcript molecule in a particular cell.
+To demonstrate the use of the `BumpyMatrix` class, let's mock up some data for a multiplexed RNA fluorescence _in situ_ hybridization (RNA-FISH) experiment.
+Our data consists of a `DataFrame` (Bioconductor's wrapper around the base `data.frame` class) where each row contains the coordinates of a single detected transcript molecule.
+We assume that demultiplexing and cell segmentation have already been performed, such that each detected molecule is also annotated with its gene identity and its cell of origin.
 
@@
 
-We use the `splitAsBumpyMatrix()` function to create a `BumpyDataFrameMatrix` where the rows are genes and the columns are cells.
-Each entry of this `BumpyDataFrameMatrix` is itself a `DataFrame` that corresponds to a particular gene/cell combination,
-and containing the x/y-coordinates for all transcript molecules of that gene in that cell.
+We use the `splitAsBumpyMatrix()` function to create a `BumpyDataFrameMatrix`, a 2-dimensional matrix abstraction where the rows are genes and the columns are cells.
+Each entry of this `BumpyDataFrameMatrix` is itself a `DataFrame` that corresponds to a particular gene/cell combination and contains the x/y-coordinates for all transcript molecules of that gene in that cell.
 Each `DataFrame` may have zero, one or multiple rows, depending on the cell's expression of the gene.
@@
 
-We can now treat `mat` in the same manner as any other matrix-like R object.
+We can now manipulate `mat` in the same manner as any other matrix-like R object.
 For example, we can slice by row and column, equivalent to filtering our original `df` by gene or cell respectively.
@@
 # Computing the Euclidean distance of each transcript from the centroid.
-# Transpositions are necessary to ensure that the cell centroid coordinates
-# are recycled along the cells, not along the features.
+# Transpositions ensure that the cell centroid coordinates are recycled along
+# the cells, not along the features.
 squared_x <- (t(mat[,,"x"]) - x_cell)^2 
@@
 
+# Examples of other applications
+
 # References

### 2023-01-19 paper.Rmd
Added introduction.

@@
 
-# Summary
+# Introduction
 
-# Statement of need
+Many bioinformatics workflows in the R/Bioconductor ecosystem are based on a 2-dimensional data representation where one dimension corresponds to features and the other dimension corresponds to samples.
+Obvious examples of this paradigm include gene-by-sample expression matrices for transcriptome profiling experiments, peak-by-sample matrices from ChIP-seq experiments, and gene-by-cell count matrices for single-cell RNA-seq experiments.
+A more sophisticated example is the `SummarizedExperiment` class, a data structure that contains any number of assay matrices, feature- and sample-level annotations as well as any additional metadata for a dataset.
+The `SummarizedExperiment` is a popular representation for genomics data, serving as the cornerstone of a wide variety of analysis pipelines ranging from RNA sequencing, epigenomics, single-cell analyses and more.
+
+However, not all types of experimental data can be easily inserted into a `SummarizedExperiment`.
+The use of assay matrices assumes that each feature/sample combination has one experimental observation (or zero observations, if missing values are allowed in the matrix).
+It is more difficult to store data types where each feature/sample combination is associated with multiple observations.
+Our motivating example is that of a highly multiplexed RNA fluorescence _in situ_ hybridization (RNA-FISH) experiment.
+Here, the features are genes, the samples are cells, and each gene/cell combination is associated with spatial coordinates for zero, one or multiple transcript molecules.
+
+The RNA-FISH transcript coordinate data has no obvious location inside the `SummarizedExperiment`.
+We cannot store it in the assays as it cannot be represented as an ordinary R matrix. 
+We might consider grouping the coordinates by cell and storing them in the sample-level annotations, but this would mean that any feature-level slicing of the `SummarizedExperiment` would not be propagated to the transcript coordinates.
+(And vice versa for storing them in the feature-level annotations, whereupon they would not respond to any sample-level slicing.)
+This negates one of the major advantages of the `SummarizedExperiment` concept - that any operations applied to a `SummarizedExperiment` will propagate across all assays and annotations contained within,
+ensuring synchronization and avoiding common book-keeping errors throughout the course of an analysis.
+
+Here, we describe the `BumpyMatrix` class, implemented in the Bioconductor package of the same name.
+This class implements a 2-dimensional array where each entry is a non-scalar object of the same type (e.g., atomic vectors, data frames) in contrast to ordinary R matrices where each entry is an atomic scalar.
+An instance of a `BumpyMatrix` class can then be stored as an assay of a `SummarizedExperiment` for synchronized manipulation of the entire dataset.
+We demonstrate the use of the `BumpyMatrix` for storing and manipulating some mock RNA-FISH data.
+We also discuss some other bioinformatics applications that can benefit from the `BumpyMatrix` concept.
 
@@
 
-To demonstrate the use of the `BumpyMatrix` class, let's mock up some data for a multiplexed RNA fluorescence _in situ_ hybridization (RNA-FISH) experiment.
+To demonstrate the use of the `BumpyMatrix` class, let's mock up some data for a multiplexed RNA-FISH experiment.
 Our data consists of a `DataFrame` (Bioconductor's wrapper around the base `data.frame` class) where each row contains the coordinates of a single detected transcript molecule.
@@
 Of particular interest is the ability to store `mat` inside Bioconductor's `SummarizedExperiment` data structure.
-The `SummarizedExperiment` provides a holistic representation of a dataset that contains all experimental assays, feature- and sample-level annotations as well as additional metadata. 
-This representation is convenient as it ensures that any operations are synchronised across all aspects of the dataset, avoiding any account-keeping errors across the lifetime of an analysis workflow.
-By storing the `BumpyDataFrameMatrix` inside the `SummarizedExperiment`, we can synchronise the manipulations of the transcript coordinates to the rest of the dataset.
+This allows us to synchronise the manipulations of the transcript coordinates to the rest of the dataset.
 

### 2023-01-19 paper.Rmd
Added some basic references.

@@
 Obvious examples of this paradigm include gene-by-sample expression matrices for transcriptome profiling experiments, peak-by-sample matrices from ChIP-seq experiments, and gene-by-cell count matrices for single-cell RNA-seq experiments.
-A more sophisticated example is the `SummarizedExperiment` class, a data structure that contains any number of assay matrices, feature- and sample-level annotations as well as any additional metadata for a dataset.
+A more sophisticated example is the `SummarizedExperiment` class [@huber2015orchestrating], a data structure that contains any number of assay matrices, feature- and sample-level annotations as well as any additional metadata for a dataset.
 The `SummarizedExperiment` is a popular representation for genomics data, serving as the cornerstone of a wide variety of analysis pipelines ranging from RNA sequencing, epigenomics, single-cell analyses and more.
@@
 It is more difficult to store data types where each feature/sample combination is associated with multiple observations.
-Our motivating example is that of a highly multiplexed RNA fluorescence _in situ_ hybridization (RNA-FISH) experiment.
+Our motivating example is that of a highly multiplexed RNA fluorescence _in situ_ hybridization (RNA-FISH) experiment [@moffitt2016highthroughput;@eng2019transcriptome]. 
 Here, the features are genes, the samples are cells, and each gene/cell combination is associated with spatial coordinates for zero, one or multiple transcript molecules.
@@
 This negates one of the major advantages of the `SummarizedExperiment` concept - that any operations applied to a `SummarizedExperiment` will propagate across all assays and annotations contained within,
-ensuring synchronization and avoiding common book-keeping errors throughout the course of an analysis.
+ensuring synchronization and avoiding common book-keeping errors throughout the course of an analysis, e.g., using the single-cell workflow described by @amezquita2020orchestrating.
 

### 2023-01-19 paper.Rmd
Added examples of other applications, more references, ack's.

@@
 
-However, not all types of experimental data can be easily inserted into a `SummarizedExperiment`.
-The use of assay matrices assumes that each feature/sample combination has one experimental observation (or zero observations, if missing values are allowed in the matrix).
+Not all types of experimental data can be easily inserted into a `SummarizedExperiment`.
+The use of assay matrices typically assumes that each feature/sample combination has one experimental observation (or zero observations, if missing values are allowed in the matrix).
 It is more difficult to store data types where each feature/sample combination is associated with multiple observations.
-Our motivating example is that of a highly multiplexed RNA fluorescence _in situ_ hybridization (RNA-FISH) experiment [@moffitt2016highthroughput;@eng2019transcriptome]. 
+An excellent example is the data generated from a highly multiplexed RNA fluorescence _in situ_ hybridization (RNA-FISH) experiment [@moffitt2016highthroughput;@eng2019transcriptome]. 
 Here, the features are genes, the samples are cells, and each gene/cell combination is associated with spatial coordinates for zero, one or multiple transcript molecules.
+We would like to hold onto these coordinates throughout the analysis as they are potentially useful in downstream steps, e.g., for visualization, diagnostics or investigation of subcellular localization.
 
-The RNA-FISH transcript coordinate data has no obvious location inside the `SummarizedExperiment`.
+Unfortunately, the RNA-FISH transcript coordinate data has no obvious location inside the `SummarizedExperiment`.
 We cannot store it in the assays as it cannot be represented as an ordinary R matrix. 
@@
 This negates one of the major advantages of the `SummarizedExperiment` concept - that any operations applied to a `SummarizedExperiment` will propagate across all assays and annotations contained within,
-ensuring synchronization and avoiding common book-keeping errors throughout the course of an analysis, e.g., using the single-cell workflow described by @amezquita2020orchestrating.
+ensuring synchronization and avoiding common book-keeping errors throughout the course of an analysis such as the single-cell workflow described by @amezquita2020orchestrating.
 
-Here, we describe the `BumpyMatrix` class, implemented in the Bioconductor package of the same name.
+Here, we describe the `BumpyMatrix` class, implemented in the Bioconductor package of the same name [@bumpymatrix].
 This class implements a 2-dimensional array where each entry is a non-scalar object of the same type (e.g., atomic vectors, data frames) in contrast to ordinary R matrices where each entry is an atomic scalar.
@@
 This compatibility also simplifies the consumption of `BumpyDataFrameMatrix` assays by frameworks that operate on a `SummarizedExperiment`.
-For example, the **alabaster** framework provides a mechanism for serializing R/Bioconductor objects into language-agnostic file formats, for use with other programming ecosystems like Python or Javascript.
+For example, the **alabaster** framework [@alabaster] provides a mechanism for serializing R/Bioconductor objects into language-agnostic file formats, for use with other programming ecosystems like Python or Javascript.
 As **alabaster** already provides a mechanism to serialize `SummarizedExperiment` objects to file, it can be easily extended to handle objects containing `BumpyDataFrameMatrix` objects in its assays.
@@
 ```{r}
-library(alabaster.se) # TODO: change to the alabaster umbrella.
+library(alabaster)
 
@@
 
-While we have largely focused on the `BumpyDataFrameMatrix`, the _BumpyMatrix_ package also supports other `BumpyMatrix` subclasses.
+While we have largely focused on the `BumpyDataFrameMatrix`, the `r Biocpkg("BumpyMatrix")` package also supports other `BumpyMatrix` subclasses.
 For example, the `BumpyNumericMatrix` class implements a matrix-like object where each entry is a numeric vector of any length.
@@
 
+# Implementation details
+
+The `BumpyMatrix` classes are based on the `CompressedList` structure from the `r Biocpkg("IRanges")` package [@lawrence2013software].
+This structure is effectively defined as a list of contiguous views into a "vector-like object" -
+for example, the `SplitDataFrameList` is a list of views into a single `DataFrame` where each view consists of a subset of rows.
+Similarly, the `CompressedNumericList` consists of a single numeric vector where each view defines a contiguous slice along the vector.
+The use of a single vector-like object enables efficient vectorization when performing operations on the `CompressedList`.
+The `BumpyMatrix` can subsequently be considered a 2-dimensional extension of the `CompressedList`,
+providing the expected methods for matrix operations, e.g., slicing by rows and columns.
+
+In principle, the `BumpyMatrix` concept can be extended to any "vector-like" object.
+This refers to any object that has a concept of its own length and contains a sequence of extractable elements.
+For example, the `GRanges` class complies with the expectations of an abstract vector,
+so one could imagine representing a two-dimensional array of genomic ranges by creating a `BumpyGRangesMatrix` class.
+In practice, any such implementation would require the definition of an existing `CompressedList` of the desired vector-like class.
+(Fortunately, the `GRangesList` class is already implemented as a `CompressedList` of `GRanges`, so a `BumpyGRangesMatrix` could build on top of that.)
+
+The `BumpyMatrix` class implements a transparent optimization when the data is "sparse", i.e., some of the vector-like objects in the entries have zero length.
+In a naive implementation, the underlying `CompressedList` would have to store a 0-length view for those entries.
+This may consume too much memory for a very large sparse matrix given that each view needs memory to hold its start and end positions.
+Instead, the `BumpyMatrix` only keeps track of its non-zero-length entries,
+in a manner similar to the compressed sparse column matrices from the `r CRANpkg("Matrix")` package [@matrix].
+This allows the creation of very large `BumpyMatrix` objects where data is only present in a small proportion of entries.
+
+Incidentally, the "bumpy" name is inspired by the 3-dimensional visualization of a `BumpyMatrix` structure.
+Image that one dimension consists of the rows, another dimension consists of the columns, and the length of the non-scalar objects projects into the third dimension.
+As the lengths are variable, this creates a bumpy surface along the plane defined by the first two dimensions.
+One can also imagine this as a 3-dimensional extension of a ragged array,
+though we did not use "ragged" so as to avoid confusion with the (rather different) `r Biocpkg("RaggedExperiment")` package.
+
 # Examples of other applications
 
+We use the `BumpyMatrix` concept for a variety of in-house applications beyond spatial transcriptomics.
+Much of this is driven by the fact that our internal databases and workflows are heavily focused on the `SummarizedExperiment` paradigm.
+Thus, it is convenient to store as much data as possible inside a `SummarizedExperiment`, for maximum compatibility with our existing data infrastructure.
+
+One example of our `SummarizedExperiment`-centric approach is our handling of drug response data across different cell lines.
+In this case, each drug is a "feature" and each cell line is a sample, and the `SummarizedExperiment` is used to store matrices of IC~50~, E~max~ and more advanced metrics [@hafner2016growth] for each drug/cell line combination.
+This representation is convenient as it allows us to easily cross-reference the columns with other `SummarizedExperiment` objects containing genotyping and expression information for the same cell lines.
+The drug response metrics are derived from raw data - namely, the cell count across a range of drug concentrations and time points.
+We also want to store the raw data in the `SummarizedExperiment` as it is useful for plotting and diagnostics during interactive examination of the dataset.
+This is easily accommodated as a `BumpyDataFrameMatrix` assay where each entry is a `DataFrame` containing the cell count data at each time and dosage.
+
+We also use the `BumpyDataFrameMatrix` to store immune repertoire sequencing datasets.
+Here, each column is a sample (or cell, for single-cell repertoire sequencing experiments) and each feature is a locus of interest, e.g., IGH, TCRA.
+Each data frame stores all detected sequences of that locus along with associated information from the AIRR standard [@vander2018airr].
+This representation is convenient as it allows us to store the repertoire data in a `MultiAssayExperiment` [@ramos2017software],
+where the samples can be cross-referenced to other `SummarizedExperiment` objects containing, e.g., expression data for the same samples.
+
+# Acknowledgements
+
+Thanks to Marc Hafner, whose problem with the representation of drug response data motivated the development of the `BumpyMatrix` in the first place.
+
 # References
